Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM

### 2013-03-15 bwamem.tex
put figures in the base directory, not in eval/

\begin{abstract}
\section{Summary:} BWA-MEM is a new alignment algorithm for aligning sequence
reads or long query sequences against a large reference genome such as human.
It automatically chooses between local and end-to-end alignments, supports
paired-end reads and performs split alignment. The algorithm is robust to
sequencing errors and applicable to a wide range of sequence lengths from 70bp
to a few megabases. For short-read mapping, BWA-MEM shows better performance
than many state-of-art read aligners to date.

@@

\subsubsection{Chaining and chain filtering} We call a group of seeds that are
colinear and close to each other as a \emph{chain}. We chain the seeds while
seeding. After that, we filter out short chains that are largely contained in a
long chain and are much worse than the long chain (by default, both 50\% and
38bp shorter than the long chain). Chain filtering aims to reduce unsuccessful
seed extension at a later step. Each chain may not always correspond to a final
hit. Chains detected here do not need to be accurate.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.4\textwidth]{eval/alnroc-se}\\
\includegraphics[width=0.4\textwidth]{eval/alnroc-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. $10^6$ pairs of 101bp reads are simulated
from the human reference genome using wgsim (http://bit.ly/wgsim2) with 1.5\%
substitution errors and 0.2\% indel variants. The insert size follows a normal
distribution $N(500,50^2)$. The reads are aligned back to the genome either as
single end (SE; top panel) or as paired end (PE; bottom panel). GEM is
configured to allow up to 5 gaps and to output suboptimal alignments (option
`--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). Its mapping quality
is estimated with a BWA-like algorithm. Other mappers are run essentially with
the default setting except for specifying the insert size distribution. The CPU
time in seconds is reported in the parantheses.}\label{fig:eval}
\end{figure}.

@@

To evaluate the performance for even longer query sequences, we aligned the
{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
the 536 strain (AC:NC\_008253) with both BWA-MEM and
nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
giving 105,505 substitution differences between the strains. BWA-MEM is slower.
It finished the alignment in 131 seconds, reporting 104,321 substitutions of
which 102,241 overlapping with the nucmer output. Manual examination of the
substitutions unique to each aligner reveals that most of them fall in short
regions with very high divergence. It is unclear which aligner is better. Note
that although Nucmer is faster in the evaluation, only BWA-MEM scales well to
large genomes.

>>>>

\begin{abstract}
\section{Summary:} BWA-MEM is a new alignment algorithm for aligning sequence
reads or long query sequences against a large reference genome such as human.
It automatically chooses between local and end-to-end alignments, supports
paired-end reads and performs split alignment. The algorithm is robust to
sequencing errors and applicable to a wide range of sequence lengths from 70bp
to a few megabases. For short-read mapping, BWA-MEM shows better performance
than several state-of-art read aligners to date.

@@

\subsubsection{Chaining and chain filtering} We call a group of seeds that are
colinear and close to each other as a \emph{chain}. We chain the seeds while
seeding. After chains are collected, we filter out short chains that are
largely contained in a long chain and are much worse than the long chain (by
default, both 50\% and 38bp shorter than the long chain). Chain filtering aims
to reduce unsuccessful seed extension at a later step. Each chain may not
always correspond to a final hit. Chains detected here do not need to be
accurate.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.4\textwidth]{alnroc-se}\\
\includegraphics[width=0.4\textwidth]{alnroc-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. $10^6$ pairs of 101bp reads are simulated
from the human reference genome using wgsim (http://bit.ly/wgsim2) with 1.5\%
substitution errors and 0.2\% indel variants. The insert size follows a normal
distribution $N(500,50^2)$. The reads are aligned back to the genome either as
single end (SE; top panel) or as paired end (PE; bottom panel). GEM is
configured to allow up to 5 gaps and to output suboptimal alignments (option
`--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). GEM does not compute
mapping quality. Its mapping quality is estimated with a BWA-like algorithm
with suboptimal alignments available. Other mappers are run essentially with
the default setting except for specifying the insert size distribution. The CPU
time in seconds is shown in the parantheses.}\label{fig:eval}
\end{figure}.

@@

To evaluate the performance for even longer query sequences, we aligned the
{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
the 536 strain (AC:NC\_008253) with both BWA-MEM and
nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
giving 105,505 substitution differences between the strains. BWA-MEM is slower.
It finished the alignment in 131 seconds, reporting 104,321 substitutions of
which 102,241 overlapping with the nucmer output. Manual examination of the
substitutions unique to each aligner reveals that most of them fall in short
regions with very high divergence. It is unclear which aligner is better. Note
that although nucmer is faster in the evaluation, only BWA-MEM scales well to
large genomes.
